From wikipedia article on housing bubble March 7

Haber, Bob. "Canadian Real Estate Bubble Blowing Up North." Forbes, Forbes Magazine, April 3, 2018, www.forbes.com/sites/bobhaber/2018/04/02/canadian-real-estate-bubble-blowing-up-north/#1b74d3871d5e.
Tencer, Daniel (October 3, 2018). Canada At Risk As 'First Cracks' Appear In Global Housing Bubbles: UBS., HuffPost (Canada edition)
"So Where Is the Next House Price Bubble Brewing?". Bloomberg.com. July 30, 2019. Archived from the original on December 14, 2019. Retrieved June 25, 2021.
"World's Bubbliest Housing Markets Flash 2008 Style Warnings". Bloomberg.com. June 15, 2021. Archived from the original on June 15, 2021. Retrieved June 25, 2021.
"Speculation by Canadians 'absolutely' playing a role in red-hot home prices: expert - National | Globalnews.ca". Global News. Retrieved April 27, 2022.
Castaldo, Joe. "How Canada's Real Estate Market Went Completely Insane." Canadian Business - Your Source For Business News, July 10, 2017, www.canadianbusiness.com/economy/how-canadas-real-estate-market-went-completely-insane/.
Andrews, Jeff. "Canada's Housing Bubble Is Starting to Burst." Curbed, Curbed, March 7, 2018, www.curbed.com/2018/3/7/17085794/canada-housing-market-collapse.
Tencer, Daniel. "Canada At Risk As 'First Cracks' Appear In Global Housing Bubbles: UBS." HuffPost Canada, HuffPost Canada, October 3, 2018, www.huffingtonpost.ca/2018/09/29/toronto-vancouver-have-world-s-3rd-and-4th-largest-housing-bubbles-ubs_a_23544956/.
"Ontario's Fair Housing Plan." News.ontario.ca, news.ontario.ca/mof/en/2017/04/ontarios-fair-housing-plan.html.
"Canada's Housing Market Still 'Highly Vulnerable' despite Easing Prices, CMHC Warns." Financial Post, October 25, 2018, business.financialpost.com/real-estate/prices-easing-but-canadas-housing-market-still-highly-vulnerable-cmhc.
Alini, Erica. "Will It Crash? Here's What to Expect from the Canadian Housing Market in 2019." Global News, Global News, December 2, 2018, globalnews.ca/news/4688308/canada-housing-market-outlook-2019/.
Khan, Mikael; Webley, Taylor (April 2019). "Disentangling the Factors Driving Housing Resales" (PDF). Bank of Canada. p. 8. Retrieved May 13, 2019.
"Investors account for a fifth of home purchases in Canada. Are they driving up housing prices in a booming market?". Retrieved June 23, 2021.
"Toronto condo, apartment rental prices drop again amid ongoing coronavirus pandemic". Global News.
"Canada Tried To Stop Real Estate Prices From Falling, And Created A Bigger Bubble". Better Dwelling. March 5, 2021.
"CMHC Warns Canadians: "Support For Homeownership Cannot Be Unlimited"". Better Dwelling. May 20, 2020.
"Bank of Canada Governor Tiff Macklem sees "some signs of excess exuberance"" – via www.youtube.com.
"BOC Governor Says Canada Will Lean On Real Estate Because "We Need The Growth"". Better Dwelling. February 25, 2021.
"Bank of Canada Wants A Housing Bubble, While Other Central Banks Try To Pop Them". Better Dwelling. March 9, 2021.
"Bank of Canada governor strongly hints that benchmark rate will stay for now". Radio Canada International. February 24, 2021.
https://money.usnews.com/investing/news/articles/2021-03-08/bank-of-canada-expected-to-resist-investor-expectations-of-early-rate-hike[bare URL]
Smith, Fergal (June 20, 2021). "Bank of Canada to break sequence of lower terminal rates as governments splurge". CTVNews. Retrieved June 25, 2021.
"Best communities in Canada: Why Atlantic Canada comes out on top". April 8, 2021.
"Rising interest rates will be 'No. 1 issue' for Canada's housing market, economists say". Global News. Retrieved June 25, 2021.
"Canada Housing Market Crash | Fast-rising Borrowing Costs Predicted to Deepen Slump".
Hogue, Robert (February 15, 2023). "Home prices still dropping across Canada". RBC Thought Leadership. Retrieved February 22, 2023.
https://edition.cnn.com/2023/01/01/business/canada-bans-home-purchases-foreigners/index.html
"How immigration and an aging population will affect Canada's housing market". financialpost. Retrieved February 22, 2023.
Barcelo, Yan. "Canadians: Don't Rush To Buy a House". Morningstar CA. Retrieved February 22, 2023.
"What $500,000 buys in today's Canadian real estate market". vancouversun. Retrieved July 18, 2021.
"Liberals need a plan to tackle housing - or they risk alienating young Canadians". Newsrooms. May 25, 2021. Retrieved May 25, 2021.
McNutt, Lydia (December 1, 2020). "Canadian Housing Market Outlook (2021)".
"Vancouver housing prices stay hot even as market cools: Royal LePage report". vancouversun. Retrieved July 18, 2021.
"Cullen Commission". cullencommission.ca. Retrieved December 20, 2020.
"Vancouver model for money laundering unprecedented in Canada, B.C. inquiry hears". CBC. Retrieved December 20, 2020.
"Canadian Real Estate Still Opaque, But Light is Creeping In". Transparency International Canada. Retrieved December 20, 2020.
"Billions In Toronto Real Estate Bought Anonymously, With Funds of Unknown Origin". Better Dwelling. March 21, 2019. Retrieved December 20, 2020.
Carmichael, Kevin (February 1, 2021). "While the world is in the midst of a tech revolution, Canadians bet on real estate". Financial Post. Retrieved April 27, 2022.
Evans, Pete (June 18, 2021). "Canadian mortgage debt grew by $18 billion in April, biggest monthly gain ever, StatsCan says". CBC News. Archived from the original on June 18, 2021.

    Kartashova, Katya; Zhou, Xiaoqing. "How Do Mortgage Rate Resets Affect Consumer Spending and Debt Repayment? Evidence from Canadian Consumers" (PDF). Bank of Canada. Retrieved July 18, 2021.

    vte

Financial bubbles
Categories:

    Economic history of CanadaReal estate bubbles of the 2010sEconomy of CanadaImmigration to CanadaMoney launderingHousing in CanadaInflation in CanadaInfrastructure in CanadaReal estate in CanadaImpact of the COVID-19 pandemic in Canada

