\documentclass[11pt]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb, amsmath}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{The Speculative motive in Housing Demand}
\author{David Robinson, Kirsten Wright}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
Source :specuiation-model.tex
\section{variables}


\begin{table}[htp]
\caption{default}
\begin{center}
\begin{tabular}{cp{8cm}}
r& global return on capital\\
U(h,x)  & individual utility for deriving consistent demand\\
 h	 &quantity of housing services for consumption \\
x     	&commodity consumption\\
$p$ 	& rental cost of a unit of housing for one period. \\%We will treat this as the $\frac{1}{T}$ of the capitalized expected stream of rent payments\\
 1			&price of the composite good, $x$\\ 
 r			& Return on capital, bank rate of interest available to a household\\
$P_{0}$ 		& Price to purchase a unit of housing at time $0$ \\
 $\kappa$ & the costs of housing transactions.\\
 $\delta_t$& discount factor for period t\\
T 			& remaining life\\
 $P_T-P_0$	&Price appreciation on the interval from 0 to $T$.\\
$\phi^s<1$	&Sellers share of sale price after transaction costs\\
$\phi^b>1$	&buyers cost with transaction costs\\
 $M\leq mP_0h$ 	&the size of mortgage, where $m\leq 0.8$\\
%	&the present value of mortgage payments to $T$.  Maybe $0.8M(1-\frac{1}{(1+r)^{T+1}})$\\
$C^C $&carrying cost of a mortgage. \\
$C^T$&Transaction cost cost of of ownership \\
$r^h$&rate of  return on the housing investment\\	
 % $P$& the path of housing prices\\
  Y & the path of annual income\\
$W_0$& Wealth at the beginning of the period (inheritance)\\
W&asset wealth, consisting of inheritance, $W_0$, minus mortgage, rate $M$,  plus sale price of home,  $P_Th$  ??? \\
%  E& the expected path of prices\\
%  &asset value of \\
%  & \\
\end{tabular}
\end{center}
\label{default}
\end{table}%

We are considering market in which one good is desired  both for its use-value, $h$, and as a financial asset\footnote{ In many advanced economies, the average annual real return on housing assets between 1950 and 2015 lies between 5 percent and 8 percent, comparable in magnitude to that of equity investment but with a lower standard deviation (Jord\'a and others 2017) From IMF’s Global Financial Stability Report – April 2018}.   We need to work out the return on housing as an asset for a homeowner. We can treat the rental price, p, as the cost of the services from a unit it of housing. The market price of a unit of housing  should then be the present value of the housing services plus the present value of the expected price appreciation. 

If a market was in equilibrium, the rate of return on a unit of housing should equal the rate of return on other assets. There would be no financial advantage to home ownership. There are some apparent advantages, including  tax exemption for capital gains on housing.\footnote{These exemptions hhave a significant cost for the government, and affect consumer behaviour. For a GE study of  the US system see Bank of Canada Working Paper 2013-33, September 2013, ``Housing and Tax Policy,'' by Sami Alpanda and Sarah Zubairy,}\footnote{Gervais (2002) considers a general-equilibrium life-cycle US economy with heterogeneous individuals, where agents can either own or rent a house. He finds that individuals at all income levels would rather live in a world where imputed rents are taxed, or one where mortgage interest payments are not deductible, and both policies have very small distributional effects in the long run}

%Start by ignoring discounting.

We need a function to calculate the return on housing.  Price appreciation on a unit of housing is $P_T-P_0$. Transaction costs including real estate fees, take a fraction from the value of the final sale, leaving $\phi^sP_Th$ where $\phi^s<1$ is the share remaining for the seller. Moving costs, legal fees and land registration costs add to the price of purchase, so that the full cost of purchase is $\phi^bP_0h$, where $\phi^b>1$

The net gain from appreciation with transaction costs is something like \[G^N=\phi^s P_T-\delta^T\phi^b P_0=P_0(\phi^s \frac{P_T}{P_0}-\delta^T\phi^b)\]%=P_0\psi(T)\]
where $\delta^T$ is the discount factor for $T$ periods. The expression $(\phi^s \frac{P_T}{P_0}-\delta^T\phi^b)=C^T$ summarizes transaction costs. $\phi^s$ and $\phi^b$ represent  fixed costs that  encourage longer ownership terms, decrease mobility and make the use of the housing stock less efficient. 

If  owners take on the maximum mortgage, $mP_0h$ and pay interest at rate $r$, %we can set the  undiscounted carrying cost of ownership is simply\footnote{
and do not pay down the mortgage the value of the infinite sum of payments $ rP_0$  minus the long tail of payments after T discounted is 
\[C^C= rM\left(\frac{1}{r} - \frac{1}{r(1+r)^T}\right)= mP_0\left(1- \frac{1}{(1+r)^T}\right) \]%=\Gamma(r,T)\]
%\[C^O= rTP_0\]
The return on the investment net of carrying costs is 
\begin{eqnarray}
R&=&G^N-C^C \\
%&=&P_0h\left(C^T-\right)\\
&=&P_0\left[ \left(\phi^s \frac{P_T}{P_0}-\delta^T\phi^b\right)      -m\left(1- \frac{1}{(1+r)^T}\right) \right]
\end{eqnarray}


The \textbf{effective rate of  return on the housing investment}, $r^h$ satisfies 
\[\left[ \left(\phi^s \frac{P_T}{P_0}-\delta^T\phi^b\right)      -m\left(1- \frac{1}{(1+r)^T}\right) \right]=(1+r^h)^T\]
%or
\[r^h=\left[ \left(\phi^s \frac{P_T}{P_0}-\delta^T\phi^b\right)      -m\left(1- \frac{1}{(1+r)^T}\right) \right]^{1/T}-1\]
and housing is a good financial investment if $r<r^h$,, i.e., id
\[r< \left[ \left(\phi^s \frac{P_T}{P_0}-\delta^T\phi^b\right)      -m\left(1- \frac{1}{(1+r)^T}\right) \right]^{1/T}-1\]

%When will it be the case that 
%\[r<\left[\phi^s \frac{P_T}{P_0}-\phi^b - \frac{1}{r}\left( 1- \frac{1}{r(1+r)^T}\right)\right]^{1/T}?\]
Writing $\frac{P_T}{P_0}= 1+r^p$, where $r^p$ ifs the compounding rate of growth of housing prices over the period $0-T$,  

\[r^p>\frac{r^T+ \frac{1}{r(1+r)^T}+\phi^b}{\phi^s}-1\]


\end{document}  