% T I T L E   P A G E
% -------------------
\pagestyle{empty}
\pagenumbering{roman}

\begin{titlepage}
        \begin{center}
        \vspace*{1.0cm}

        \Huge
        {\bf Financialization of the Housing Market: A Contribution to Modern Urban Rent Theory}

        \vspace*{1.0cm}

        \normalsize
        by \\

        \vspace*{1.0cm}

        \Large
        Kirsten Wright \\

        \vspace*{3.0cm}

        \normalsize
        A thesis \\
        presented to the University of Waterloo \\ 
        in fulfillment of the \\
        thesis requirement for the degree of \\
        Doctor of Philosophy \\
        in \\
        Systems Design Engineering \\

        \vspace*{2.0cm}

        Waterloo, Ontario, Canada, 2024 \\

        \vspace*{1.0cm}

        \copyright\ Kirsten Wright 2024 \\
        \end{center}
\end{titlepage}

% The rest of the front pages should contain no headers and be numbered using Roman numerals starting with `ii'
\pagestyle{plain}
\setcounter{page}{2}

\cleardoublepage % Ends the current page and causes all figures and tables that have so far appeared in the input to be printed.
% In a two-sided printing style, it also makes the next page a right-hand (odd-numbered) page, producing a blank page if necessary.
\phantomsection    % allows hyperref to link to the correct page
 
% E X A M I N I N G   C O M M I T T E E (Required for Ph.D. theses only)
% Remove or comment out the lines below to remove this page
\addcontentsline{toc}{chapter}{Examining Committee}
\begin{center}\textbf{Examining Committee Membership}\end{center}
  \noindent
The following served on the Examining Committee for this thesis. The decision of the Examining Committee is by majority vote.
  \bigskip
  
  \noindent
\begin{tabbing}
Internal-External Member: \=  \kill % using longest text to define tab length
External Examiner: \>  Naoki Yoshihara \\ 
\> Associate Professor, Department of Economics, \\
\> University of Massachusetts Amherst \\ % Toronto \\
\end{tabbing} 
  \bigskip
  
  \noindent
\begin{tabbing}
Internal-External Member: \=  \kill % using longest text to define tab length
Supervisor(s): \> Jangho Yang \\
\> Assistant Professor, Department of Management Science and \\ 
\> Engineering, University of Waterloo \\ \\
\> Sean Geobey \\
\> Associate Professor, School of Environment, Enterprise and \\
\> Development, University of Waterloo \\
\end{tabbing}
  \bigskip

  \noindent
\begin{tabbing}
Internal-External Member: \=  \kill % using longest text to define tab length
Internal Member: \> Chrystopher Nehaniv \\
\> Professor, Department of Systems Design Engineering, \\
\> University of Waterloo \\
\end{tabbing}
  \bigskip

  \noindent
\begin{tabbing}
Internal-External Member: \=  \kill % using longest text to define tab length
Internal Member: \> Terry Stewart \\
\> Adjunct Professor, Department of Psychology, \\
\> University of Waterloo \\
\end{tabbing}
  \bigskip

  \noindent
  \begin{tabbing}
Internal-External Member: \=  Frances Westley \\ % \kill % using longest text to define tab length
% Internal Member: \> \\
\> Professor, School of Environment, Enterprise and Development, \\
\> University of Waterloo \\
\end{tabbing}
  \bigskip


  
%   \noindent
% \begin{tabbing}
% Internal-External Member: \=  \kill % using longest text to define tab length
% Other Member(s): \> Supervisor Name \\
% \> Professor, Dept. of Systems Design Engineering, University of Waterloo \\
% \end{tabbing}

\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% D E C L A R A T I O N   P A G E
% -------------------------------
  % The following is a sample Declaration Page as provided by the GSO
  % December 13th, 2006.  It is designed for an electronic thesis.
 \addcontentsline{toc}{chapter}{Author's Declaration}
 \begin{center}\textbf{Author's Declaration}\end{center}
  
 \noindent
I hereby declare that I am the sole author of this thesis. This is a true copy of the thesis, including any required final revisions, as accepted by my examiners.

  \bigskip
  
  \noindent
I understand that my thesis may be made electronically available to the public.

\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% A B S T R A C T
% ---------------
\addcontentsline{toc}{chapter}{Abstract}

\input{00-abstract}

\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% A C K N O W L E D G E M E N T S
% -------------------------------
\addcontentsline{toc}{chapter}{Acknowledgements}
\begin{center}\textbf{Acknowledgements}\end{center}

I don't know how to begin this acknowledgement. When I began this I didn't have my two children. It feels like a different world. Although there have been more challenges than could have expected, it is hard to imagine a more engaging or stimulating project to get to work on. 

I feel enormous gratitude to Waterloo for the chance it offered me to be part of this community.  My first job was at Blackberry, in Rim 4, on the floor just above Mike Lazaridis, and next to the early engineering team, early on I got the chance on clocks for the Long Now Foundation, sculptures with Philip Beesley and Rob Gorbet, underwater robots, light based communication protocols, and reverse engineering the Canada arm.

Then I had the good fortune of getting to join Frances Westley's team working on social innovation and participating in the community with The Waterloo Institute for Complex Systems (WICI). WICI  and associated institutions, gathered  people like Lee Smolin, Brian Arthur, Roberto Unger, Doyne Farmer, Jaron Lanier and so many others to work on how our understanding of complex systems translates to the transformation of social and ecological systems. 

For this thesis, I feel like everyone in my family did this single handedly except me. My husband Jason by insisting I go away over march break even though he had the stomach flu, was working full-time, and watching our two little kids, my sister who let me stay at her house and helped edit every chapter, my mother and father who insisted I come away for a three focused weeks to just get a draft out, fed me and took me snowshoeing every day, and most especially to my father who talk through all of the economics with, helped me work through the math, and understand pieces, over the course of years. My daughter Jackie, now 6, knows this. I said had to go away to work on finishing. She patted me and said `at least you have your sister and your dad to help you.' She is right.

I also want to thank my committee. Thank you Frances and Terry for staying a part of this as it's evolved from early work on data assimilation and resilience in agent based models, to work on specific models of housing markets and rent. Thank you to Chrystopher for joining in the last year, and for the inspiration of your work on currencies and climate. Thank you Sean for being a collaborator for so long. Thank you Jangho. I never expected to find a supervisor whose work engaged with so many of the pieces of this work, let alone to find one in the Engineering Faculty at Waterloo. You bring together a background in engineering and a PhD in classical economics, with recent work with some of the most exciting groups working on  complex systems. It has been wonderful to get the chance to work with you over the last two and a half years. I  hope I get the chance to collaborate with all of you in the future. 

If there is an organizing logic for this work, it is the tagline for the systems design design department: 'systems is how we understand the world and design is how we change it.' I'm grateful to be part of so many communities engaging with these questions in different ways. Thank you.

\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% D E D I C A T I O N
% -------------------
\addcontentsline{toc}{chapter}{Dedication}
\begin{center}\textbf{Dedication}\end{center}

\begin{center}
To Wyatt, Jackie, and Jason.    
\end{center}

\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% T A B L E   O F   C O N T E N T S
% ---------------------------------
\renewcommand\contentsname{Table of Contents}
\tableofcontents
\cleardoublepage
\phantomsection    % allows hyperref to link to the correct page

% L I S T   O F   F I G U R E S
% -----------------------------
\addcontentsline{toc}{chapter}{List of Figures}
{\renewcommand{\addvspace}[1]{} \listoffigures}
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page

% L I S T   O F   T A B L E S
% ---------------------------
\addcontentsline{toc}{chapter}{List of Tables}
{\renewcommand{\addvspace}[1]{} \listoftables}
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page

% L I S T   O F   A B B R E V I A T I O N S
% ---------------------------
\renewcommand*{\abbreviationsname}{List of Abbreviations}
\printglossary[type=abbreviations]
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page

% L I S T   O F   S Y M B O L S
% ---------------------------
\printglossary[type=symbols]
\cleardoublepage
\phantomsection		% allows hyperref to link to the correct page


% Change page numbering back to Arabic numerals
\pagenumbering{arabic}

