
\documentclass[a4paper]{article}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{amssymb}
\usepackage{enumerate}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[english]{babel}
\usepackage[autostyle]{csquotes}
% \usepackage[backend=biber,style=authoryear]{biblatex}
% \addbibresource{jmp.bib}
\usepackage[normalem]{ulem}
\title{Interest Rates}
\author{ }
\date{\today}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{assumption}{Assumption}
\newtheorem{remark}{Remark}
\newtheorem{condition}{Condition}
\newtheorem{proposition}{Proposition}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\begin{document}
\maketitle
\begin{abstract}
This is  the abstract.
\end{abstract}

\section{Introduction}

If people are investing their own money, high interest rates attract them to investment. If people are investing other people's money high interest lower borrow, while low interest rates encourage them to borrow more, even if they have to invent new ways to push it into less productive activities. % invent ways to shove it into other less productive activities. / invent unproductive ways to 


---


David Graeber said that when they lowered interest rates they hoped people would borrow lots and lots of money and invest it productive uses. Instead people borrowed lots and lots of money and buried it their backyards. It got worse after he died. More recently they've buried it at the heart of cities displacing both people and productive uses.

Piketty has identified this gap between the real growth rate, which is the marginal productivity of capital in Keynsian economics, and the marginal rate of return. 
And arguing that in many cases the return on capital is higher than the return on production.


there is a separation between the return on capital, the marginal return on capital which could correspond to the growth rate of the cost of capital, which is using other people moneys.


---


it's an open system


DiR you gotta understand the bank of Canada's job is to scare and choke. The problem for it now is while they can scare, the only people they can choke are the innocents, workers, and homeowners.

They currently have a mighty urge to choice. 
 However they can't choke the war mongers, oil, shipping, international food supplies, 

Open and closed systems.

% \bibliographystyle{plainnat} % or try abbrvnat or unsrtnat
% \bibliography{jmp.bib} % refers to example.bib
% % \printbibliography
% % javascript:void(0);
\end{document}